The Forecasting Trap -- notes

### 2022-02-11 notebook/alternate-models.Rmd
pretty close

@@

```{r}
# these dynamics are nonsense
# predator pop goes to Inf if prey is extinct
# prey = sp 1, predator= sp 2
R1 <- R2 <- 3
A1 <- A2 <- 2
B1 <- 0.5
B2 <- -0.3
Tmax <- 50
omega1 <- rnorm(Tmax, 0, 0.001)
omega2 <- rnorm(Tmax, 0, 0.001)

u1 <- u2 <- s1 <- s2 <- x1 <- x2 <- numeric(Tmax)
x1[1] <- 1
x2[1] <- 1

# or, constant strategy
u1 <- rep(.2, Tmax)
u2 <- rep(.5, Tmax)

for (t in 1:(Tmax-1)) {
    s1[t] <- x1[t] * (1 - u1[t])
    s2[t] <- x2[t] * (1 - u2[t])
    x1[t+1] <- s1[t] * exp(R1 - A1 * s1[t] - B1 * s2[t] + omega1[t])
    x2[t+1] <- s2[t] * exp(R2 - A2 * s1[t] - B2 * s1[t] + omega2[t])
}

df <- data.frame(time = 1:Tmax, x1, x2, u1, u2)

```{r}
ggplot(df) +
 geom_path(aes(time, x1)) + 
 geom_path(aes(time, x2), col="red")
```

@@

model_5sp <- function(u_prey = 0.45, u_pred = .66) {
    Tmax <- 50
    x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
    s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
    xi1 <- rnorm(Tmax, 0, 0.01)
    xi2 <- rnorm(Tmax, 0, 0.01)
    xi3 <- rnorm(Tmax, 0, 0.01)
    xi4 <- rnorm(Tmax, 0, 0.01)
    xi5 <- rnorm(Tmax, 0, 0.01)

    x1[1] <- x2[1] <- x3[1] <- 1
    x4[1] <- x5[1] <- .5

@@

    }
    df <- tibble(t = 1:Tmax, x1, x2, x3, x4, x5)
    df
}

df <- model_5sp()

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "population") %>%
    ggplot(aes(t, population, col= species)) + geom_path()
```

```{r}
clip <- function(x, lower=0, upper=1) {
    x[x <= lower] <- lower
    x[x >= upper] <- upper
    x
}

utility <- function(x) {
    delta <- 0.99
    w1 <- 0.25
    w2 <- 0.25
    w3 <- 1 - (w1 + w2)
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    df <- model_5sp(u_prey, u_pred)
    R <- 
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * x5
    t <- seq_along(R)
    - as.numeric(R %*% delta ^ t)
    }

-utility(c(0.5,0.5))
-utility(c(0,1))
-utility(c(1,0))

o <- optim(c(0.5,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
o$par
o$value

o <- optim(c(0.5,0.5), utility, method="SANN")
clip(o$par, 0, 1)
o$value

@@

- Determine n alternate models over the same state space. Consider structurally different models, ideally with structures consistent with
  common forecast / inference models

>>>>

- Model 1: A reasonable approximation model that does not produce very accurate forecasts, but nevertheless leads to good management decisions.
- Model 2: A reasonable approximation model that produces more accurate forecasts but leads to substantially worse management outcomes
- Model 3: The "True" Model: five-species model; functional forms and parameter values are never disclosed to the 'manager' but are used to simulate all data and management scenarios.

@@

```{r}
## helper funs
format_sim <- function(df) {
  df %>%  
  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
  pivot_longer(any_of(c("cormorant", "bass", "herring")),
                  names_to = "species",
                  values_to = "biomass")
}
```

@@

```{r model_5sp}
# x1,x2,x3 Herring (Prey)
# x4 Bass (predator)
# x5 Cormorant (conservation target, predator)
model_5sp <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.01,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50,
                      reps = 40) {
    df <- tibble()
    for(r in 1:reps){
      x1 <- x2 <- x3 <- x4 <- x5 <- numeric(Tmax)
      s1 <- s2 <- s3 <- s4 <- numeric(Tmax)
      xi1 <- rnorm(Tmax, 0, sigma)
      xi2 <- rnorm(Tmax, 0, sigma)
      xi3 <- rnorm(Tmax, 0, sigma)
      xi4 <- rnorm(Tmax, 0, sigma)
      xi5 <- rnorm(Tmax, 0, sigma)

      x1[1] <- x1o
      x2[1] <- x2o
      x3[1] <- x3o
      x4[1] <- x4o
      x5[1] <- x5o

@@

      }
      df <- dplyr::bind_rows(tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r), df)
    }
    df
}

@@

```{r utility}
clip <- function(x, lower=0, upper=1) {
    x[x <= lower] <- lower
    x[x >= upper] <- upper
    x
}

## Compute the reward associated with a given simulation df given 
## rates for prey and predator species (x[1],x[2])
reward <- function(df,
                   x,
                   w3 = 0.5,              # Conservation value
                   w2 = 0.25,             # Predator harvest value
                   w1 = 1 - (w2+w3),      # Prey harvest value
                   delta = 0.999
                  ) {
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    R <-
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * df$x5
    t <- seq_along(R)
    as.numeric(R %*% delta ^ t)
}

# Function to be optimized for fixed choice of harvest rates x[1],x[2]
utility <- function(x, f = model_5sp, ...) {

    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    df <- f(u_prey, u_pred, ...)

    # average accross the reps    
    df <- df %>% group_by(time) %>% summarise(across(.fns=mean))

    # negative since optimizer minimizes
    - reward(df, x)
}

o <- optim(c(0.01,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
o$par
o$value

o <- optim(c(0.01,0.5), utility)
clip(o$par, 0, 1)
o$value

x <- c(0.0,.58)
utility(x)

For the sake of argument, let us allow the model of Brias & Munch to represent the "true" model, which is unknown.
As researchers, we only have access to data derived from observations from the model. 
Let us consider a sample simulation of such data:

```{r}
df <- model_5sp(0, .0, sigma = 0.01)

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + geom_path() + ggtitle("Unfished")
```

Starting from the unfished equilibrium,

```{r}
df <- model_5sp(0, .58, sigma = 0.01)

df %>%
    pivot_longer(tidyselect::starts_with("x"),
                    names_to = "species",
                    values_to = "biomass") %>%
    ggplot(aes(time, biomass, group= interaction(species,r), col=species)) + 
  geom_path() + ggtitle("Optimally managed")
```

There are many other models, including process-based models, statistical models or even black-box machine learning models
which 5 dimensional state-space and two-dimensional action space.

```{r}
## Three-species 'view' of the underlying dynamics, without fishing, starting all species from a low level

#true_observations <- model_5sp(0, 0., sigma = 0.05, x1o =.033, x2o = .033, x3o = .033, x4o = .1, x5o = .1, Tmax = 50)
true_observations <- model_5sp(0, 0, sigma = 0.05)

obs <- true_observations %>% 
  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass")

obs %>%
    ggplot(aes(time, biomass, col = species, group = interaction(species,r))) +
    geom_path(alpha=0.4)
```

Now consider an alternative model which ignores some of this biological complexity:
Our alternate model has several oversimplifications:
(1) it treat all three herring species as a single species,
(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
and ignoring the impact of cormorant's predation on the herring itself.
(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.

The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the unfished dynamics.
Under any harvesting regime or other influence that perturbs the system significantly far from the unfished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.

```{r}
model_3sp <- function(u_prey = 0,
                      u_predator = 0,
                      A1 = 2,
                      R1 = 1.6,
                      A2 = 2.75,
                      R2 = 1.45,
                      Tmax = 50,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      sigma = 0.01,
                      reps=40) {
    df <- tibble()
    for(r in 1:reps){
    omega1 <- rnorm(Tmax, 0, sigma)
    omega2 <- rnorm(Tmax, 0, sigma)
    omega3 <- rnorm(Tmax, 0, sigma)

    s1 <- x1 <- numeric(Tmax)
    s4 <- x4 <- numeric(Tmax)
    x2 <- x3 <- numeric(Tmax)
    x5 <- numeric(Tmax)

    x1[1] <- x1o + x2o + x3o
    x4[1] <- x4o
    x5[1] <- x5o

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - 10*u_prey)
        s4[t] <- x4[t] * (1 - u_predator)

        x1[t+1] <- s1[t] * exp(R1 - s1[t] * A1 + omega1[t])
        x4[t+1] <- s4[t] * exp(R2 - s4[t] * A2 + omega2[t])
        x5[t+1] <- .25 * x1[t]
    }

    df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
    }
  df
}
```

Does this model produce good forecasts?

```{r}
# unfished equilibrium
m1_forecast <- model_3sp(0, 0.0, sigma = 0.05,  Tmax = 50)

predictions <- m1_forecast %>%
  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
  pivot_longer(any_of(c("cormorant", "bass", "herring")),
                  names_to = "species",
                  values_to = "biomass") %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predictions %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
  geom_point(data = filter(obs, r==1))
```

```{r}
# unfished equilibrium
m1_forecast <- model_3sp(0, 0.58, sigma = 0.05,  Tmax = 50)

predictions <-   
m1_forecast %>%
    transmute(time = time, cormorant = x5, bass = x4, herring = x1, r=r) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass") %>% 
    group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predictions %>%
ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
  geom_point(data = filter(obs, r==1))
```

```{r}
# Optimal policy under identical utility
u_3sp <- function(x) utility(x, f = model_3sp)

m3sp <- optim(c(0.5,0.1), u_3sp)
m3sp$value
clip(m3sp$par, 0, 1)

Alternate model, also mistakenly treats the three species of herring as a single species, but does a better job capturing the predator-prey relationships.  This model also happens to produce better forecasts, but leads to worse decision:

```{r}
model_B <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.05,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50,
                      R1 =  1.15, #1.0213,
                      A11 = 1,
                      A14 = 0.1, # 0.7252,
                      A15 = 0.1, # 0.2445,
                      A41 =  0.8, #0.7252,
                      A51 = 0.3, # 0.2445,
                    reps = 40) {

    df <- tibble()
    for(r in 1:reps){

      x1 <- x4 <- x5 <- numeric(Tmax)
      s1 <-  s4 <- numeric(Tmax)
      x2 <- x3 <- numeric(Tmax)
      xi1 <- rnorm(Tmax, 0, sigma)
      xi4 <- rnorm(Tmax, 0, sigma)
      xi5 <- rnorm(Tmax, 0, sigma)

      x1[1] <- x1o + x2o + x3o
      x4[1] <- x4o
      x5[1] <- x5o

      for (t in 1:(Tmax-1)) {
          s1[t] <- x1[t] * (1 - u_prey)
          s4[t] <- x4[t] * (1 - u_pred)
          x1[t+1] = s1[t]*exp(R1 - A11*s1[t] - A14*s4[t] - A15 * x5[t] + xi1[t])
          x4[t+1] = s4[t]*exp(A41*s1[t] - s4[t] + xi4[t])
          x5[t+1] = x5[t]*exp(A51*s1[t] - x5[t] +  xi5[t])

      }
      df <- bind_rows(df, tibble(time = 1:Tmax, x1, x2, x3, x4, x5, r=r))
    }
    df
}
```

```{r}
# Optimal policy under identical utility
u_B <- function(x) utility(x, f = model_B)
u_B(c(0, .58))
#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)

B <- optim(c(0.5,0.1), u_B)
B$value
model_B_policy <- clip(B$par, 0, 1)
model_B_policy

Run the simulation using management policy of each

Tmax <- 20
x <- model_B_policy
#x <- c(0.,0.58)  # Following the optimal policy under model_B

obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predict_stats_B <- predict_B %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predict_stats_B %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))

```{r}
## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
Tmax <- 5
x <- c(0,0.58)

obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predict_stats <- predict %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

## dotted are forecasted values, solid are observed values

predict_stats %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))

```{r}
score_me <- function(x = c(0,0.58), process=model_5sp, forecast=model3p, Tmax=5) {

  obs <- process(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
  predict <- forecast(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

  #scoringRules::crps_sample(predict$x1, )

}
